---
title: "DE analysis"
author: "Sergey Naumenko"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

# Overview
- Principal Investigator: Beth Overmoyer
- Experiment: RNAseq_analysis_of_inflammatory_breast_cancer_hbc04141
- study 6 was excluded because if low read depth in 3373-3
- https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html
- [AnnotationHub](https://bioconductor.org/packages/release/bioc/html/AnnotationHub.html). We use ensembl version matching bcbio pipeline - v94.
- [HBC materials](https://hbctraining.github.io/DGE_workshop_salmon/lessons/03_DGE_QC_analysis.html)
- [HBC materials - functional analysis](https://hbctraining.github.io/DGE_workshop_salmon/lessons/functional_analysis_2019.html)
- http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

```{r include = TRUE}
## Setup
### Bioconductor and CRAN libraries used

library(DESeq2)
library(tidyverse)
library(RColorBrewer)
library(pheatmap)
library(DEGreport)
library(tximport)
library(ggplot2)
library(ggrepel)
library(knitr)
library(AnnotationHub)
library(ensembldb)

ggplot2::theme_set(theme_light(base_size = 14))

opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)
```
```{r include = TRUE}
# Have a folder called `data`, and copy your Salmon folders here from the cluster. 
## List all directories containing data  
### change the pattern to something specific to your Salmon folders
samples <- list.files(path = "./data/final", 
                      full.names = T,
                      pattern = "^S")

## Obtain a vector of all filenames including the path
files <- file.path(samples, "salmon", "quant.sf")
files

## Since all quant files have the same name it is useful to have names for each element
### change the string in str_replace so the pattern matches your filenames
names(files) <- str_replace(samples, "./data/final/", "") 
```

```{r include = TRUE}
# Load the data and metadata
meta <- read_csv("tables/metadata_corrected.csv") %>% 
  column_to_rownames(var = "samplename")
protein_coding_genes <- read_csv("tables/ensembl_w_description.protein_coding.csv")
```
```{r include = TRUE}
# Connect to AnnotationHub
ah <- AnnotationHub()

# Query AnnotationHub
hs_ens <- query(ah, c("Homo sapiens", "EnsDb"))

# Get Ensembl94 - used in bcbio
hs_ens <- hs_ens[["AH64923"]]

# Extract gene-level information
txdb <- transcripts(hs_ens,
                    return.type = "data.frame") %>% 
  dplyr::select(tx_id, gene_id)  

genedb <- genes(hs_ens, 
                return.type = "data.frame") %>% 
  dplyr::select(gene_id, gene_name, symbol)  

gene_symbol <- genedb %>% dplyr::select(gene_id, symbol)

hsdb <- inner_join(txdb, genedb)
write.table(hsdb, 
            file = "data/ensembl94_hg38_annotations.txt", 
            sep = "\t", 
            row.names = F, 
            quote = F)
```
```{r include = TRUE}
# Read in  a tx2gene file with transcript identifiers in the first column and gene identifiers in the second column
#wormdb <- read.table("ensembl94_WBcel235_annotations.txt", sep="\t", header=T)
tx2gene <- hsdb[, c("tx_id", "gene_id")]

# Run tximport
txi_file <- "data/txi.RDS"
if (file.exists(txi_file)){
    txi <- readRDS(txi_file)
}else{
    txi <- tximport(files, 
                type = "salmon", 
                tx2gene = tx2gene, 
                countsFromAbundance = "lengthScaledTPM", 
                ignoreTxVersion = FALSE)
    saveRDS(txi, txi_file)
}

# Look at the counts
class(txi)
attributes(txi)
txi$counts %>% View()
```
# Checking to see that the transcript to gene mapping is correct
When you have annotations that are from a different source from your reference you can run into problems (i.e lose genes). Some checks you can do before proceeding:

1. Look at the dimensions of your count matrix. Do you have ~20k genes present? `dim(txi$counts)`
2. When running `tximport()` you will get a message in your console. If you see something like `transcripts missing from tx2gene` start troubleshooting.
```{r}
dim(txi$counts)
```

# Sanity check that metadata matches your expression
It is always a good idea to check if:

1. Do you have expression data for all samples listed in your metadata?
2. Are the samples in your expression data in the same order as your metadata?
```{r}
### Check that sample names match in both files
all(colnames(txi$counts) %in% rownames(meta))

# Not the same? Make them the same
### This will change depending on what names you have listed!
#paste0(meta$samplename, "_", meta$library)
#rownames(meta) <- paste0(meta$samplename, "_", meta$library)
#meta$genotype <- relevel(meta$genotype, ref="Wildtype")

### Check that sample names match in both files
all(colnames(txi$counts) %in% rownames(meta))

### Check that all samples are in the same order
meta <- meta[colnames(txi$counts),]
all(colnames(txi$counts) == rownames(meta))

```

# Run DESeq2
```
estimating size factors
estimating dispersions
gene-wise dispersion estimates
mean-dispersion relationship
final dispersion estimates
fitting model and testing
```

* [Estimating size factors and count normalization](https://hbctraining.github.io/DGE_workshop_salmon/lessons/02_DGE_count_normalization.html)
* [Gene-wise dispersions](https://hbctraining.github.io/DGE_workshop_salmon/lessons/04_DGE_DESeq2_analysis.html)
* [Mean-dispersion(variance) relationship and the Negative Binomial Model](https://hbctraining.github.io/DGE_workshop_salmon/lessons/01_DGE_setup_and_overview.html#modeling-count-data)
* [Model fitting and hypothesis testing](https://hbctraining.github.io/DGE_workshop_salmon/lessons/05_DGE_DESeq2_analysis2.html)


# Wald test
*Here we subset protein coding genes*.
```{r include = TRUE}
## Create DESeq2Dataset object
dds_file <- "data/dds.RDS"
meta$treatment <- as.factor(meta$treatment)
meta$response <- as.factor(meta$response)
meta$er <- as.factor(meta$er)
meta$date_of <- as.factor(meta$date_of)
meta$tumor_percentage <- as.factor(meta$tumor_percentage)
meta$tumor_percentage_high <- as.factor(meta$tumor_percentage_high)

non_responders <- meta %>% dplyr::filter(study_id %in% c(2, 19)) %>% row.names() 

if (file.exists(dds_file)){
    dds <- readRDS(dds_file)
}else{
    dds <- DESeqDataSetFromTximport(txi,
                                colData = meta, 
                                design = ~treatment)
    
    dds <- dds[,!colnames(dds) %in% non_responders]
    design(dds) <- formula(~treatment + response + er + tumor_percentage_high + date_of)
    
    # subset protein-coding genes
    pc_genes <- intersect(protein_coding_genes$ensembl_gene_id, row.names(dds))
    dds <- dds[pc_genes,]
    # 100 reads / 40 samples
    keep <- rowSums(counts(dds)) >= 100
    dds <- dds[keep,]

    # Run DESeq2
    dds <- DESeq(dds)
    saveRDS(dds, dds_file)
}
```

# DEGreport QC

## Size factor QC - samples 1-15
```{r fig.width=10, fig.height = 10}
counts <- counts(dds, normalized = TRUE)
design <- as.data.frame(colData(dds))
degCheckFactors(counts[, 1:5])

```

## Size factor QC - samples 16-30
```{r fig.width=10, fig.height = 10}
degCheckFactors(counts[, 16:30])

```

## Size factor QC - samples 31-40
```{r fig.width=10, fig.height = 10}
degCheckFactors(counts[, 31:40])

```

# Mean-Variance QC plots

## treatment
```{r fig.width=15, fig.height = 15}
res <- results(dds)
degQC(counts, design[["treatment"]], pvalue = res[["pvalue"]])
```

## response
```{r fig.width=15, fig.height = 15}
degQC(counts, design[["response"]], pvalue = res[["pvalue"]])
```

## ER
```{r fig.width=15, fig.height = 15}
degQC(counts, design[["er"]], pvalue = res[["pvalue"]])
```

## tumor_percentage_high
```{r fig.width=15, fig.height = 15}
degQC(counts, design[["tumor_percentage_high"]], pvalue = res[["pvalue"]])
```

# Covariates effect on count data
```{r}
mdata <- colData(dds) %>%  as.data.frame()  %>% 
  dplyr::select(treatment, response, er, date_of, tumor_percentage_high)
resCov <- degCovariates(log2(counts(dds)+0.5), mdata)
```

```{r}
mdata %>% ggplot(aes(tumor_percentage_high, fill = response)) + geom_bar(position = "dodge2")
```

# Covariates correlation with metrics
```{r}
cor <- degCorCov(mdata)
```
```{r}
mdata %>% ggplot(aes(tumor_percentage_high, fill = treatment)) + geom_bar(position = "dodge2")
```


```{r}
mdata %>% ggplot(aes(date_of, fill = response)) + geom_bar(position = "dodge2")
```

# Sample-level QC analysis 

```{r include = TRUE}
### Transform counts for data visualization (unsupervised analysis)
rld_file <- "data/rld.RDS"
if (file.exists(rld_file)){
    rld <- readRDS(rld_file)
}else{
    rld <- rlog(dds, blind = TRUE)
    saveRDS(rld, rld_file)
}
class(rld) # what type of object is this

# we also need just a matrix of transformed counts
rld_mat <- assay(rld)
```

## PCA - treatment
```{r}
# Use the DESeq2 function
plotPCA(rld, intgroup = c("treatment"))
```

## PCA - response
```{r}
# Use the DESeq2 function
plotPCA(rld, intgroup = c("response"))
```

## PCA - ER
```{r}
# Use the DESeq2 function
plotPCA(rld, intgroup = c("er"))
```

## PCA - tumor_percentage
```{r}
# Use the DESeq2 function
plotPCA(rld, intgroup = c("tumor_percentage"))
```

## PCA - tumor_percentage_high
```{r}
# Use the DESeq2 function
plotPCA(rld, intgroup = c("tumor_percentage_high"))
```

## PCA - date_of
```{r}
# Use the DESeq2 function
plotPCA(rld, intgroup = c("date_of"))
```

# Inter-correlation analysis

## Without study_id

```{r, fig.width = 20, fig.height = 20}
# Correlation matrix
rld_cor <- cor(rld_mat)

meta$study_id <- as.factor(meta$study_id)
# Create annotation file for samples
annotation <- meta[, c("treatment", "response", "er", "tumor_percentage_high", "date_of")]

# Change colors
heat.colors <- brewer.pal(6, "Blues")

# Plot heatmap
pheatmap(rld_cor, 
         annotation = annotation, 
         border = NA,
         fontsize = 20)
```

## With study_id

```{r, fig.width = 20, fig.height = 20}
# Correlation matrix
rld_cor <- cor(rld_mat)

meta$study_id <- as.factor(meta$study_id)
# Create annotation file for samples
annotation <- meta[, c("treatment", "response", "er", "tumor_percentage_high", "date_of", "study_id")]

# Change colors
heat.colors <- brewer.pal(6, "Blues")

# Plot heatmap
pheatmap(rld_cor, 
         annotation = annotation, 
         border = NA,
         fontsize = 20)
```

# Treatment Post vs Pre - see Table3
```{r}
# Get results for treatment
contrast <- c("treatment", "post", "pre")
resTreatment <- results(dds, contrast = contrast, alpha = 0.05)
length(which(resTreatment$padj < 0.05))

# Add annotations
resTreatment_tb <- resTreatment %>%
  data.frame() %>% 
  rownames_to_column(var = "gene") %>% 
  as_tibble() %>% 
  left_join(gene_symbol, by = c("gene" = "gene_id"))

resTreatment_tb_significant <- dplyr::filter(resTreatment_tb, padj < 0.05)

samples_pre <- meta %>% dplyr::filter(treatment == "pre") %>% row.names() 

counts_pre <- txi$abundance %>% 
  as.data.frame() %>% 
  dplyr::select(one_of(samples_pre)) %>% 
  rowMeans() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "ensembl_gene_id") 
  
colnames(counts_pre) <- c("ensembl_gene_id", "pre_expression_mean_tpm")

samples_post <- meta %>% dplyr::filter(treatment == "post") %>% row.names() 

counts_post <- txi$abundance %>% 
        as.data.frame() %>% 
        dplyr::select(one_of(samples_post)) %>%
        rowMeans() %>% 
        as.data.frame() %>% 
        rownames_to_column(var = "ensembl_gene_id")
colnames(counts_post) <- c("ensembl_gene_id", "post_expression_mean_tpm")

counts_post <- counts_post %>% left_join(counts_pre, by = c("ensembl_gene_id" = "ensembl_gene_id"))

resTreatment_tb_significant <- resTreatment_tb_significant %>% 
          left_join(counts_post, by = c("gene" = "ensembl_gene_id"))

write_csv(resTreatment_tb_significant, 
          "tables/T3.DE_treatment.csv")

# Separate into up and down-regulated gene sets
sigTreatment_up <- rownames(resTreatment)[which(resTreatment$padj < 0.01 & resTreatment$log2FoldChange > 0)]
sigTreatment_down <- rownames(resTreatment)[which(resTreatment$padj < 0.01 & resTreatment$log2FoldChange < 0)]
```

# Response Yes vs No - see Table4
```{r}
# Get results for rescue vs wt
contrast <- c("response", "Yes", "No")
resResponse <- results(dds, contrast = contrast, alpha = 0.05)
length(which(resResponse$padj < 0.05))

# Add annotations
resResponse_tb <- resResponse %>%
  data.frame() %>% 
  rownames_to_column(var = "gene") %>% 
  as_tibble() %>% 
  left_join(gene_symbol, by = c("gene" = "gene_id")) 

resResponse_tb_significant <- dplyr::filter(resResponse_tb, padj < 0.05)

samples_no <- meta %>% dplyr::filter(response == "No") %>% row.names() 

counts_no <- txi$abundance %>%
                as.data.frame() %>% 
                dplyr::select(any_of(samples_no)) %>% 
                rowSums() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")
colnames(counts_no) <- c("ensembl_gene_id", "no_expression_mean_tpm")

samples_yes <- meta %>% dplyr::filter(response == "Yes") %>% row.names() 

counts_yes <- txi$abundance %>% 
                as.data.frame() %>% 
                dplyr::select(any_of(samples_yes)) %>% 
                rowSums() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")

colnames(counts_yes) <- c("ensembl_gene_id", "yes_expression_mean_tpm")

counts_yes <-counts_yes %>% 
            left_join(counts_no, 
                      by = c("ensembl_gene_id" = "ensembl_gene_id"))

resResponse_tb_significant <- resResponse_tb_significant %>% 
          left_join(counts_yes, by = c("gene" = "ensembl_gene_id"))


write_csv(resResponse_tb_significant, 
          "tables/T4.DE_response.csv")

# Separate into up and down-regulated gene sets
sigResponse_up <- rownames(resResponse)[which(resResponse$padj < 0.01 & resResponse$log2FoldChange > 0)]
sigResponse_down <- rownames(resResponse)[which(resResponse$padj < 0.01 & resResponse$log2FoldChange < 0)]
```

# ER : Positive vs Negative - Table 5
```{r}
contrast <- c("er", "Positive", "Negative")
resER <- results(dds, contrast = contrast, alpha = 0.05)
length(which(resER$padj < 0.05))

# Add annotations
resER_tb <- resER %>%
  data.frame() %>% 
  rownames_to_column(var = "gene") %>% 
  as_tibble() %>% 
  left_join(gene_symbol, by = c("gene" = "gene_id")) 

resER_tb_significant <- dplyr::filter(resER_tb, padj < 0.05)

samples_pos <- meta %>% dplyr::filter(er == "Positive") %>% row.names() 

counts_pos <- txi$abundance %>%
                as.data.frame() %>% 
                dplyr::select(any_of(samples_pos)) %>% 
                rowMeans() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")
colnames(counts_pos) <- c("ensembl_gene_id", "Positive_expression_mean_tpm")

samples_neg <- meta %>% dplyr::filter(er == "Negative") %>% row.names() 

counts_neg <- txi$abundance %>% 
                as.data.frame() %>% 
                dplyr::select(any_of(samples_neg)) %>% 
                rowMeans() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")

colnames(counts_neg) <- c("ensembl_gene_id", "Negative_expression_mean_tpm")

counts_pos <-counts_pos %>%
            left_join(counts_neg, 
                      by = c("ensembl_gene_id" = "ensembl_gene_id"))

resER_tb_significant <- resER_tb_significant %>%
          left_join(counts_pos, by = c("gene" = "ensembl_gene_id"))


write_csv(resER_tb_significant,
          "tables/T5.DE_ER.csv")

# Separate into up and down-regulated gene sets
sigER_up <- rownames(resER)[which(resER$padj < 0.01 & resER$log2FoldChange > 0)]
sigER_down <- rownames(resER)[which(resER$padj < 0.01 & resER$log2FoldChange < 0)]
```

# tumor_percentage_high : High vs Low - Table 6
```{r}
contrast <- c("tumor_percentage_high", "high", "low")
resTP <- results(dds, contrast = contrast, alpha = 0.05)
length(which(resTP$padj < 0.05))

# Add annotations
resTP_tb <- resTP %>%
  data.frame() %>% 
  rownames_to_column(var = "gene") %>% 
  as_tibble() %>% 
  left_join(gene_symbol, by = c("gene" = "gene_id")) 

resTP_tb_significant <- dplyr::filter(resTP_tb, padj < 0.05)

samples_high <- meta %>% dplyr::filter(tumor_percentage_high == "high") %>% row.names() 

counts_high <- txi$abundance %>%
                as.data.frame() %>% 
                dplyr::select(any_of(samples_high)) %>% 
                rowMeans() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")
colnames(counts_high) <- c("ensembl_gene_id", "High_expression_mean_tpm")

samples_low <- meta %>% dplyr::filter(tumor_percentage_high == "low") %>% row.names() 

counts_low <- txi$abundance %>% 
                as.data.frame() %>% 
                dplyr::select(any_of(samples_low)) %>% 
                rowMeans() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")

colnames(counts_low) <- c("ensembl_gene_id", "Low_expression_mean_tpm")

counts_high <-counts_high %>%
            left_join(counts_low, 
                      by = c("ensembl_gene_id" = "ensembl_gene_id"))

resTP_tb_significant <- resTP_tb_significant %>%
          left_join(counts_high, by = c("gene" = "ensembl_gene_id"))


write_csv(resTP_tb_significant,
          "tables/T6.DE_tumor_percentage_high.csv")

# Separate into up and down-regulated gene sets
sigTP_up <- rownames(resTP)[which(resTP$padj < 0.01 & resTP$log2FoldChange > 0)]
sigTP_down <- rownames(resTP)[which(resTP$padj < 0.01 & resTP$log2FoldChange < 0)]
```


# date_of: 20180323 vs 20180228 - Table 7
```{r}
contrast <- c("date_of", "20180323", "20180228")
resDO <- results(dds, contrast = contrast, alpha = 0.05)
length(which(resDO$padj < 0.05))

# Add annotations
resDO_tb <- resDO %>%
  data.frame() %>% 
  rownames_to_column(var = "gene") %>% 
  as_tibble() %>% 
  left_join(gene_symbol, by = c("gene" = "gene_id")) 

resDO_tb_significant <- dplyr::filter(resDO_tb, padj < 0.05)

samples_23 <- meta %>% dplyr::filter(date_of == "20180323") %>% row.names() 

counts_23 <- txi$abundance %>%
                as.data.frame() %>% 
                dplyr::select(any_of(samples_23)) %>% 
                rowMeans() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")
colnames(counts_23) <- c("ensembl_gene_id", "20180323_expression_mean_tpm")

samples_28 <- meta %>% dplyr::filter(date_of == "20180228") %>% row.names() 

counts_28 <- txi$abundance %>% 
                as.data.frame() %>% 
                dplyr::select(any_of(samples_28)) %>% 
                rowMeans() %>% 
                as.data.frame() %>% 
                rownames_to_column(var = "ensembl_gene_id")

colnames(counts_28) <- c("ensembl_gene_id", "20180228_expression_mean_tpm")

counts_23 <-counts_23 %>%
            left_join(counts_28, 
                      by = c("ensembl_gene_id" = "ensembl_gene_id"))

resDO_tb_significant <- resDO_tb_significant %>%
          left_join(counts_23, by = c("gene" = "ensembl_gene_id"))


write_csv(resDO_tb_significant,
          "tables/T7.DE_dete_of.csv")

# Separate into up and down-regulated gene sets
sigDO_up <- rownames(resDO)[which(resDO$padj < 0.01 & resDO$log2FoldChange > 0)]
sigDO_down <- rownames(resDO)[which(resDO$padj < 0.01 & resDO$log2FoldChange < 0)]
```

# Visualization

*Gene example*

```{r}
d <- plotCounts(dds, 
                gene = "ENSG00000130234", 
                intgroup = "treatment", 
                returnData = TRUE)

ggplot(d, aes(x = treatment, y = count, color = treatment)) + 
     geom_point(position = position_jitter(w = 0.1, h = 0)) +
     geom_text_repel(aes(label = rownames(d))) + 
     theme_bw(base_size = 10) +
     ggtitle("ACE2") +
     theme(plot.title = element_text(hjust = 0.5)) +
     scale_y_log10()
     
```


```{r, fig.width = 10, fig.height = 10}
# Add a column for significant genes
resTreatment_tb <- resTreatment_tb %>% mutate(threshold = padj < 0.01)

## Volcano plot
ggplot(resTreatment_tb) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("Treatment Post vs Pre") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-10,10)) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))
```

```{r}
# Add a column for significant genes
resResponse_tb <- resResponse_tb %>% mutate(threshold = padj < 0.01)

ggplot(resResponse_tb) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("Response Yes vs No") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-10,10)) +
  scale_y_continuous(limits = c(0, 7.5))+
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))
```

```{r}
# Add a column for significant genes
resER_tb <- resER_tb %>% mutate(threshold = padj < 0.01)

ggplot(resER_tb) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("ER: Positive vs Negative") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-10,10)) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))
```

```{r}
# Add a column for significant genes
resTP_tb <- resTP_tb %>% mutate(threshold = padj < 0.01)

ggplot(resTP_tb) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("Tumor_percentage_high: High vs Low") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-10,10)) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))
```

```{r}
# Add a column for significant genes
resDO_tb <- resDO_tb %>% mutate(threshold = padj < 0.01)

ggplot(resDO_tb) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("Dafe of: 20180323 vs 20180228") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-10,10)) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))
```
# Heatmaps

```{r, fig.width = 20, fig.height = 30}
# Create a matrix of normalized expression
sig_up <- resTreatment_tb_significant %>% arrange(-log2FoldChange) %>% head(50) %>% pull(gene)
sig_down <- resTreatment_tb_significant %>% arrange(log2FoldChange) %>% head(50) %>% pull(gene)
sig <- c(sig_up, sig_down)

row_annotation <- gene_symbol %>% 
                    as_tibble() %>% 
                    dplyr::filter(gene_id %in% sig)

plotmat <- txi$abundance[c(sig_up, sig_down),] %>% as.data.frame() %>% 
          rownames_to_column(var = "ensembl_gene_id") %>% 
          left_join(gene_symbol, by = c("ensembl_gene_id" = "gene_id")) %>% 
          drop_na(symbol)

plotmat$ensembl_gene_id <- NULL

plotmat <- plotmat %>% column_to_rownames(var = "symbol") %>% as.matrix()

# Color palette
heat.colors <- brewer.pal(6, "YlOrRd")

# Plot heatmap
# color = heat.colors,
pheatmap(plotmat, scale = "row", 
         show_rownames = TRUE,
         border = FALSE,
         annotation = meta[, c("treatment"), drop = FALSE],
         main = "Top 50 Up- and Down- regulated genes in treatment: post vs pre",
         fontsize = 20)

```


```{r, fig.width = 20, fig.height = 30}
# Create a matrix of normalized expression
sig_up <- resResponse_tb_significant %>% arrange(-log2FoldChange) %>% head(50) %>% pull(gene)
sig_down <- resResponse_tb_significant %>% arrange(log2FoldChange) %>% head(50) %>% pull(gene)
sig <- c(sig_up, sig_down)

row_annotation <- gene_symbol %>% 
                    as_tibble() %>% 
                    dplyr::filter(gene_id %in% sig)

plotmat <- txi$abundance[c(sig_up, sig_down),] %>% as.data.frame() %>% 
          rownames_to_column(var = "ensembl_gene_id") %>% 
          left_join(gene_symbol, by = c("ensembl_gene_id" = "gene_id")) %>% 
          drop_na(symbol)

plotmat$ensembl_gene_id <- NULL

plotmat <- plotmat %>% column_to_rownames(var = "symbol") %>% as.matrix()

# Color palette
heat.colors <- brewer.pal(6, "YlOrRd")

# Plot heatmap
pheatmap(plotmat, 
         scale = "row", 
         show_rownames = TRUE,
         border = FALSE,
         annotation = meta[, c("response"), drop = FALSE],
         main = "Top 50 Up- and Down- regulated genes in Response: Yes vs No",
         fontsize = 20)

```

```{r, fig.width = 20, fig.height = 30}
# Create a matrix of normalized expression
sig_up <- resER_tb_significant %>% arrange(-log2FoldChange) %>% head(50) %>% pull(gene)
sig_down <- resER_tb_significant %>% arrange(log2FoldChange) %>% head(50) %>% pull(gene)
sig <- c(sig_up, sig_down)

row_annotation <- gene_symbol %>% 
                    as_tibble() %>% 
                    dplyr::filter(gene_id %in% sig)

plotmat <- txi$abundance[c(sig_up, sig_down),] %>% as.data.frame() %>% 
          rownames_to_column(var = "ensembl_gene_id") %>% 
          left_join(gene_symbol, by = c("ensembl_gene_id" = "gene_id")) %>% 
          drop_na(symbol)

plotmat$ensembl_gene_id <- NULL

plotmat <- plotmat %>% column_to_rownames(var = "symbol") %>% as.matrix()

# Color palette
heat.colors <- brewer.pal(6, "YlOrRd")

# Plot heatmap
pheatmap(plotmat, 
         scale = "row", 
         show_rownames = TRUE,
         border = FALSE,
         annotation = meta[, c("er"), drop = FALSE],
         main = "Top 50 Up- and Down- regulated genes in ER: positive vs negative",
         fontsize = 20)

```



```{r, fig.width = 20, fig.height = 10}
# Create a matrix of normalized expression
sig_up <- resTP_tb_significant %>% arrange(-log2FoldChange) %>% head(50) %>% pull(gene)
sig_down <- resTP_tb_significant %>% arrange(log2FoldChange) %>% head(50) %>% pull(gene)
sig <- c(sig_up, sig_down)

row_annotation <- gene_symbol %>% 
                    as_tibble() %>% 
                    dplyr::filter(gene_id %in% sig)

plotmat <- txi$abundance[c(sig_up, sig_down),] %>% as.data.frame() %>% 
          rownames_to_column(var = "ensembl_gene_id") %>% 
          left_join(gene_symbol, by = c("ensembl_gene_id" = "gene_id")) %>% 
          drop_na(symbol)

plotmat$ensembl_gene_id <- NULL

plotmat <- plotmat %>% column_to_rownames(var = "symbol") %>% as.matrix()

# Color palette
heat.colors <- brewer.pal(6, "YlOrRd")

# Plot heatmap
pheatmap(plotmat, 
         scale = "row", 
         show_rownames = TRUE,
         border = FALSE,
         annotation = meta[, c("tumor_percentage_high"), drop = FALSE],
         main = "Top Up/Down-regulated genes in Tumor_percentage_high: high vs low",
         fontsize = 20)

```


```{r, fig.width = 20, fig.height = 30}
# Create a matrix of normalized expression
sig_up <- resDO_tb_significant %>% arrange(-log2FoldChange) %>% head(50) %>% pull(gene)
sig_down <- resDO_tb_significant %>% arrange(log2FoldChange) %>% head(50) %>% pull(gene)
sig <- c(sig_up, sig_down)

row_annotation <- gene_symbol %>% 
                    as_tibble() %>% 
                    dplyr::filter(gene_id %in% sig)

plotmat <- txi$abundance[c(sig_up, sig_down),] %>% as.data.frame() %>% 
          rownames_to_column(var = "ensembl_gene_id") %>% 
          left_join(gene_symbol, by = c("ensembl_gene_id" = "gene_id")) %>% 
          drop_na(symbol)

plotmat$ensembl_gene_id <- NULL

plotmat <- plotmat %>% column_to_rownames(var = "symbol") %>% as.matrix()

# Color palette
heat.colors <- brewer.pal(6, "YlOrRd")

# Plot heatmap
pheatmap(plotmat, 
         scale = "row", 
         show_rownames = TRUE,
         border = FALSE,
         annotation = meta[, c("response"), drop = FALSE],
         main = "Top 50 Up- and Down- regulated genes in date_of: 20180323 vs 20180228",
         fontsize = 20)

```


# R session

```{r}
sessionInfo()
```